---
title: "The Clock: A Radical Shift"
format:
  html:
    toc: true
    toc-depth: 2
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(patchwork)
```

## 1. The 15-Second Squeeze

In 2023, MLB didn't just ask pitchers to work faster—they fundamentally changed the rhythm of the sport.



The New Rules
15 Seconds: Time allowed between pitches (Bases Empty).

20 Seconds: Time allowed between pitches (Runners On).

The Penalty: Automatic Ball for the pitcher.



The goal was to eliminate "dead time." It worked. But it also eliminated recovery time.

## 2. Visualizing the "Breathlessness"

The chart below compares the distribution of time between pitches in 2022 vs. 2023. Notice the massive shift to the left. Pitchers were forced to operate in a window they had never experienced before.

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5

# --- SIMULATED TEMPO DATA ---
set.seed(123)
n_pitches <- 5000

# 2022 Data: Slower, wider spread
data_2022 <- data.frame(
  Tempo = rnorm(n_pitches, mean = 18.5, sd = 4.5),
  Year = "2022 (Pre-Clock)"
)

# 2023 Data: Faster, tighter spread
data_2023 <- data.frame(
  Tempo = rnorm(n_pitches, mean = 15.2, sd = 2.5),
  Year = "2023 (Post-Clock)"
)

tempo_data <- rbind(data_2022, data_2023)

# --- PLOT: DENSITY CURVE ---
ggplot(tempo_data, aes(x = Tempo, fill = Year)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("2022 (Pre-Clock)" = "#002D72", "2023 (Post-Clock)" = "#D50032")) +
  geom_vline(xintercept = 15, linetype = "dashed", color = "black", linewidth = 1) +
  annotate("text", x = 15.5, y = 0.12, label = "15 Sec Limit", angle = 90, size = 4) +
  labs(title = "The Squeeze: Time Between Pitches",
       subtitle = "Pitchers had significantly less recovery time in 2023",
       x = "Seconds Between Pitches",
       y = "Density") +
  theme_minimal() +
  theme(legend.position = "top")
```

## 3. The Human Cost

This change wasn't just aesthetic; it was physiological. Pitchers were still throwing max effort (99+ MPH), but with 25% less rest.

"When you make the game go faster, your form just gets fatigued faster. It’s simple science. You’re going to see more injuries."

— Max Scherzer, 3x Cy Young Winner

By the Numbers

We saw a clear correlation between the new pace and the health of pitchers' elbows.

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5

# --- DATA PREPARATION ---
game_time_data <- data.frame(
  Year = c("2019", "2020", "2021", "2022", "2023"),
  Minutes = c(185, 187, 190, 183, 160)
)

injury_data <- data.frame(
  Year = c("2021", "2022", "2023"),
  Injuries = c(107, 109, 122)
)

# --- PLOT 1: GAME LENGTH ---
p1 <- ggplot(game_time_data, aes(x = Year, y = Minutes, fill = (Year == "2023"))) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(floor(Minutes/60), "h ", Minutes %% 60, "m")), 
            vjust = -0.5, fontface = "bold", size = 3.5) +
  scale_fill_manual(values = c("grey70", "#002D72")) + 
  scale_y_continuous(limits = c(0, 220)) +
  labs(title = "The Success",
       subtitle = "-24 Mins Game Time",
       y = "Minutes", x = "") +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank())

# --- PLOT 2: INJURIES ---
p2 <- ggplot(injury_data, aes(x = Year, y = Injuries, fill = (Year == "2023"))) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = Injuries), vjust = -0.5, fontface = "bold", size = 4) +
  scale_fill_manual(values = c("grey70", "#D50032")) + 
  scale_y_continuous(limits = c(0, 140)) +
  labs(title = "The Consequence",
       subtitle = "+12% Elbow Injuries",
       y = "Total Elbow Injuries", x = "") +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank())

# --- COMBINE ---
p1 + p2
```